---
title: Analysis of Barchard's Self-Assessed Components of Emotional Intelligence
author: Darren R. Skidmore
date: 2026-04-10
bibliography: barchard_report.bib
---

```{r}
#| echo: false
#| warning: false
#| message: false

# Load libraries
library(readr)
library(psych)
library(ggplot2)
```

# Abstract

Emotional intelligence (EI) is an aspect of intelligence which concerns the abilities of understanding, utilizing, and managing one's emotions in social and personal contexts. The term "emotional intelligence" was coined fairly recently [@salovey1990] but has seen significant interest in popular culture and psychology research since then, with the term having over one million search results on Google Scholar as of 2026. Today, there are many debates about the specifics of emotional intelligence: is it itself a form of intelligence or is it a byproduct of intelligence, or is it caused by nature or nurture [@dwyer2022]. For these reasons, its measurement is much less established than IQ or academic achievement despite the development of various emotional intelligence models.

In her doctoral thesis, Kimberly Barchard of the University of British Columbia proposed one such model of emotional intelligence including 14 constructs. Barchard identified half of these subcomponents as personality dimensions, meaning they can be measured by self-assessment rather than being cognitive abilities measured by performance [@barchard2001]. For this reason, it was proposed that there is no support in considering the aggregate of the seven personality subcomponents as emotional intelligence and the seven dimensions are best interpreted independently of each other.

Items from the International Personality Item Pool (IPIP) administered to the Eugene-Springfield Community Sample were used to derive summaries such as a full reliability analysis, factor analysis, and sample-based scoring tables for the seven Barchard subcomponents. The findings show that the subcomponents are moderately to highly reliable, with sufficient sample sizes and confidence intervals to derive scale scoring tables with respect to this sample. A factor analysis demonstrated modest support for the theoretical subcomponents, with sufficient factor loadings but also a high prevalence of cross-loading. There was no support for a global composite.

# Method

## Sample

The Eugene-Springfield Community Sample was a longitudinal study conducted by the Oregon Research Institute which involved recruited adult participants in the Eugene, Oregon area completing self-administered personality assessments over a span of fourteen years. The raw data are available on the Harvard Dataverse website and consists of 1,134 cases with demographic information and 960 cases that completed at least one survey.

Many of the questions on the surveys are part of the International Personality Item Pool (IPIP), a public domain collection of 3,320 items from various personality inventories [@goldberg2006; @goldberg2018]. Sixty-eight questions from four of the inventories were identified as being related to the seven Barchard personality subcomponents [@ipip_ei_key].

Only participants with at least gender and age information available and who completed all sixty-eight questions (n=524) were used in the study. The average age of participants was 51.5 with a range of 20-85. 58% of participants were female while 42% were male. Nearly 98% of participants were white. Participants had a very high level of education with 54% having a college degree and 98% having at least a high school degree or vocational/technical training.

```{r}
#| echo: false
#| warning: false
#| message: false
#| results: asis

demographics_df <- read_csv(
  "../output/barchard/barchard_demographics.csv",
  show_col_types = FALSE
)

n_sample <- nrow(demographics_df)
gender_n <- sum(demographics_df$SEX %in% c("F", "M"), na.rm = TRUE)

fmt_pct <- function(x, total) {
  sprintf("%.1f%%", 100 * x / total)
}

count_row <- function(level, count, total) {
  data.frame(
    Value = level,
    Summary = as.character(count),
    Percentage = fmt_pct(count, total)
  )
}

education_levels <- c(
  "Less than high school",
  "High school or vocational training",
  "Some college",
  "College graduate",
  "Some post-college education",
  "Post-college degree"
)

education_codes <- c("1", "2", "3", "4", "5", "6", "7", "8")
education_labels <- c(
  "Less than high school",
  "Less than high school",
  "High school or vocational training",
  "High school or vocational training",
  "Some college",
  "College graduate",
  "Some post-college education",
  "Post-college degree"
)

education_factor <- factor(
  demographics_df$EDUC,
  levels = education_codes,
  labels = education_labels
)

ethnicity_factor <- factor(
  ifelse(demographics_df$ETHNIC == "5", "White", "Non-White / Other"),
  levels = c("White", "Non-White / Other")
)

education_n <- sum(!is.na(education_factor))
ethnicity_n <- sum(!is.na(ethnicity_factor))

age_table <- data.frame(
  Statistic = c("Mean (SD)", "Range"),
  Summary = c(
    sprintf(
      "%.1f (%.1f)",
      mean(demographics_df$AGE, na.rm = TRUE),
      sd(demographics_df$AGE, na.rm = TRUE)
    ),
    sprintf(
      "%d-%d",
      min(demographics_df$AGE, na.rm = TRUE),
      max(demographics_df$AGE, na.rm = TRUE)
    )
  )
)

gender_table <- rbind(
  count_row("Female", sum(demographics_df$SEX == "F", na.rm = TRUE), gender_n),
  count_row("Male", sum(demographics_df$SEX == "M", na.rm = TRUE), gender_n)
)

education_table <- do.call(
  rbind,
  lapply(
    education_levels,
    function(level) {
      count_row(
        level,
        sum(education_factor == level, na.rm = TRUE),
        education_n
      )
    }
  )
)

ethnicity_table <- rbind(
  count_row(
    "White",
    sum(ethnicity_factor == "White", na.rm = TRUE),
    ethnicity_n
  ),
  count_row(
    "Non-White / Other",
    sum(ethnicity_factor == "Non-White / Other", na.rm = TRUE),
    ethnicity_n
  )
)

cat("### Age\n")
knitr::kable(age_table, align = c("l", "l", "l"))

cat("\n### Gender\n")
knitr::kable(gender_table, align = c("l", "l", "l"))

cat("\n### Ethnicity\n")
knitr::kable(ethnicity_table, align = c("l", "l", "l"))

cat("\n### Education\n")
knitr::kable(education_table, align = c("l", "l", "l"))
```

## Measurement

The seven subcomponents of emotional intelligence which were identified by Barchard as being personality dimensions are as follows:

- Positive Expressivity: The tendency to express one’s positive emotions nonverbally
- Negative Expressivity: The tendency to express one’s negative emotions nonverbally
- Attending to Emotions: The tendency to attend to emotions and be aware of them
- Emotion-Based Decision-Making: The tendency to make plans and decisions based on one’s feelings rather than basing them on logic
- Responsive Distress: The tendency to become distressed when in the presence of other people who are distressed
- Responsive Joy: The tendency to become happy or cheerful when in the presence of other people who are happy or cheerful
- Empathic Concern: The tendency to feel concern or sympathy for those who suffer

The sixty-eight test items were derived from the full set of IPIP questionnaires, including 49 items from the Personality, Emotions, and Attitudes (PEA) inventory, 4 items from the 179-IPIP inventory, 7 items from the 858-IPIP inventory, and 8 items from the 275-IPIP inventory [@ori_technical_report].

Positive Expressivity and Emotion-Based Decision-Making had nine items each, while the other subcomponents had ten items each. Some items, such as "Express my affection physically" were positively related to the scale, and other items, such as "Keep my feelings to myself, regardless of how unhappy I am" were negatively related to the scale and were negative keyed. Negatively keyed items were scored by applying the function `6 - n` to its raw score (e.g., an answer of "4" would become "2".)

After items were appropriately keyed, the raw totals for the subscales were computed.

## Descriptive Statistics

```{r}
#| echo: false
#| warning: false
#| message: false

scale_descriptives <- read_csv(
  "../output/barchard/barchard_scale_descriptives.csv",
  show_col_types = FALSE
)

knitr::kable(scale_descriptives, digits = 2)
```

## Inter-Scale Correlations

Most scales had low-to-moderate positive correlations to each other (range 0.13-0.58).

```{r}
#| echo: false
#| warning: false
#| message: false

scale_correlations <- read_csv(
  "../output/barchard/barchard_scale_correlations.csv",
  show_col_types = FALSE
)

knitr::kable(scale_correlations, digits = 2)
```

## Reliability

The reliability measures Cronbach's alpha ($\alpha$) and McDonald's omega ($\omega$) for each scale were nearly identical and ranged between 0.68 (Responsive Distress) and 0.83 (Attending to Emotions), suggesting acceptable to good internal consistency. Mean inter-item correlation ranged from 0.17 (Responsive Distress) to 0.33 (Positive Expressivity).

```{r}
#| echo: false
#| warning: false
#| message: false

scale_reliabilities <- read_csv(
  "../output/barchard/barchard_scale_reliabilities.csv",
  show_col_types = FALSE
)

knitr::kable(scale_reliabilities, digits = 2)
```

# Factor Analysis

Factor analysis--an essential aspect of psychometrics--is a process used to determine the number of latent factors in a sample and which variables account for each of the factors. A factor analysis was performed to determine if there was support for arranging the data in seven factors.

## Preliminary Tests

```{r}
#| echo: false
barchard_df <- read_csv(
  "../output/barchard/barchard_scored.csv",
  show_col_types = FALSE
)
barchard_df$id <- as.character(barchard_df$id)

barchard_key <- read_csv("../measures/barchard.csv")
fa_items <- barchard_key$id
barchard_df_fa <- barchard_df[, fa_items]
```

### Bartlett's Test of Sphericity
Verified with a confidence test that the data are not an identity matrix and therefore suitable for factor analysis.

```{r}
cortest.bartlett(barchard_df_fa)
```

### Kaiser-Meyer-Olkin Test
Verified that the data are suitable for factor analysis with an overall MSA of 0.88 (excellent).
```{r}
KMO(barchard_df_fa)
```

### Parallel Analysis
Parallel analysis uses two different methods: principal component analysis (PCA) and principal axis factoring (PAF) to examine at what point the successive eigenvectors decrease below a threshold value (1).

```{r}
set.seed(1)
fa.parallel(barchard_df_fa)
```

Though parallel analysis suggests that the number of factors = 11, the factor eigenvalues experience a much smaller derivative beyond 8 factors despite still being slightly above the threshold value. Therefore, the existence of at least 7 factors is clearly supported, with factors 8 through 11 being borderline.

### Principal Axis Factoring
Principal Axis Factoring (PAF), which is a type of exploratory common factor analysis, is the preferred method for factor analysis [@kim2008].

Out of sixty-eight questions, sixty-four (94%) had at least one loading ≥ 0.30 and twenty-seven (40%) were cross-loaded, meaning they had a loading ≥ 0.30 on two or more components. Forty-nine (72%) questions had their highest loading match their dominant theoretical scale.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 18

structure_df <- read_csv(
  "../output/barchard/barchard_fa_structure.csv",
  show_col_types = FALSE
)

assignment_df <- read_csv(
  "../output/barchard/barchard_fa_structure_assignments.csv",
  show_col_types = FALSE
)

factor_cols <- grep("^PA", colnames(structure_df), value = TRUE)
row_labels <- ifelse(
  assignment_df$cross_loading,
  paste0(structure_df$id, "*"),
  structure_df$id
)
structure_mat <- as.matrix(structure_df[, factor_cols])
rownames(structure_mat) <- row_labels

heatmap_df <- as.data.frame(
  as.table(structure_mat),
  responseName = "loading"
)
colnames(heatmap_df) <- c("item_label", "factor", "loading")

heatmap_df$item_label <- factor(
  heatmap_df$item_label,
  levels = rev(row_labels)
)
heatmap_df$factor <- factor(heatmap_df$factor, levels = factor_cols)

assignment_df$item_label <- ifelse(
  assignment_df$cross_loading,
  paste0(assignment_df$id, "*"),
  assignment_df$id
)

primary_tiles <- assignment_df[
  assignment_df$primary_loading_abs >= 0.30,
  c("item_label", "primary_factor")
]
colnames(primary_tiles) <- c("item_label", "factor")

secondary_tiles <- assignment_df[
  assignment_df$secondary_loading_abs >= 0.30,
  c("item_label", "secondary_factor")
]
colnames(secondary_tiles) <- c("item_label", "factor")

primary_tiles$item_label <- factor(
  primary_tiles$item_label,
  levels = rev(row_labels)
)
primary_tiles$factor <- factor(primary_tiles$factor, levels = factor_cols)

secondary_tiles$item_label <- factor(
  secondary_tiles$item_label,
  levels = rev(row_labels)
)
secondary_tiles$factor <- factor(secondary_tiles$factor, levels = factor_cols)

ggplot(heatmap_df, aes(x = factor, y = item_label, fill = loading)) +
  geom_tile(color = "#DDDDDD", linewidth = 0.2) +
  geom_tile(
    data = primary_tiles,
    aes(x = factor, y = item_label),
    fill = NA,
    color = "black",
    linewidth = 0.6,
    inherit.aes = FALSE
  ) +
  geom_tile(
    data = secondary_tiles,
    aes(x = factor, y = item_label),
    fill = NA,
    color = "#CC7A00",
    linewidth = 0.6,
    linetype = "22",
    inherit.aes = FALSE
  ) +
  scale_fill_gradient2(
    low = "#3B4CC0",
    mid = "#F7F7F7",
    high = "#B40426",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    title = "7-Factor Structure Matrix Heatmap",
    subtitle = paste(
      "Black border = primary loading >= .30;",
      "orange dashed border = secondary loading >= .30"
    ),
    x = "Empirical Factor",
    y = "Item ID",
    fill = "Structure\nLoading",
    caption = "* item flagged as cross-loading in the structure-assignment table"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    axis.text.y = element_text(size = 6),
    axis.text.x = element_text(size = 9),
    plot.caption = element_text(hjust = 0)
  )
```

Each factor in the factor analysis was named after its most common (dominant) scale group with respect to the theoretical model. None of the new factors were exactly the same as their theoretical components. The components Attending to Emotions, Emotion-Based Decision-Making, and Empathic Concern each had 9/10 of their original items, but the other components only retained about half of their original items, with the remainder going to precisely one other scale. These relations are shown in the form `theoretical_component -> observed_component` where all `theoretical_component` items with a different highest loading instead had their highest loading with `observed_component`. 

- Positive Expressivity -> Responsive Joy
- Negative Expressivity -> Responsive Distress
- Responsive Distress -> Empathic Concern
- Emotion-Based Decision Making -> Attending to Emotions

```{r}
#| echo: false
#| warning: false
#| message: false

pfa_component_align <- read_csv(
  "../output/barchard/barchard_fa_structure_factor_alignment.csv",
  show_col_types = FALSE
)

knitr::kable(pfa_component_align, digits = 2)
```

Finally, the intercorrelations between the seven observed factors were computed. Because scale intercorrelations are weak-to-none, there is no support for a global composite score.
```{r}
#| echo: false
#| warning: false
#| message: false

pfa_component_phi <- read_csv(
  "../output/barchard/barchard_fa_phi.csv",
  show_col_types = FALSE
)

knitr::kable(pfa_component_phi, digits = 2)
```

# Conclusion

The seven personality subcomponents were supported as practical scored domains by acceptable internal consistency and low-to-moderate inter-scale correlations, although factor analysis showed only partial support for the intended item-to-scale structure. The seven-factor model was retained as the most interpretable and theory-aligned structure; although exploratory factor analysis also suggested more complex 8- to 11-factor solutions which were not considered due to the manifestation of very small composites with only 1 to 3 items.

Support for the intended scale mapping was only partial as the items in the factor analysis did not perfectly align with their theoretical subcomponents. The vast majority of the items had at least one strong loading, but a significant minority of the items cross-loaded or had two or more strong loadings on components. Some pairs of components, such as Positive Expressivity and Responsive Joy, were similar enough that some test items could fit into either.

The factor-analyzed composite and original composite intercorrelations showed no support for a global composite. This is likely related to the fact that this study used self-assessment data from only half of Barchard's emotional intelligence components, with the other components being impractical to measure with self-assessment techniques.

# Appendix A - Items

The sixty-eight (68) items are listed below.

```{r}
#| echo: false
#| warning: false
#| message: false

appendix_key <- read_csv(
  "../measures/barchard.csv",
  show_col_types = FALSE
)

appendix_assignments <- read_csv(
  "../output/barchard/barchard_fa_structure_assignments.csv",
  show_col_types = FALSE
)

appendix_alignment <- read_csv(
  "../output/barchard/barchard_fa_structure_factor_alignment.csv",
  show_col_types = FALSE
)

scale_lookup <- unique(appendix_key[, c("scale_id", "scale")])

factor_scale_lookup <- merge(
  appendix_alignment[, c("factor", "dominant_scale_id")],
  scale_lookup,
  by.x = "dominant_scale_id",
  by.y = "scale_id",
  all.x = TRUE,
  sort = FALSE
)

factor_labels <- setNames(
  factor_scale_lookup$scale,
  factor_scale_lookup$factor
)

appendix_table <- merge(
  appendix_key[, c("id", "name", "scale", "key")],
  appendix_assignments[, c(
    "id",
    "primary_factor",
    "primary_loading",
    "secondary_factor",
    "secondary_loading",
    "weak_primary",
    "cross_loading"
  )],
  by = "id",
  all.x = TRUE,
  sort = FALSE
)

appendix_table$primary_factor <- unname(
  factor_labels[appendix_table$primary_factor]
)
appendix_table$secondary_factor <- unname(
  factor_labels[appendix_table$secondary_factor]
)

appendix_table$weak_primary <- ifelse(
  appendix_table$weak_primary,
  "\u2713",
  ""
)
appendix_table$cross_loading <- ifelse(
  appendix_table$cross_loading,
  "\u2713",
  ""
)

colnames(appendix_table) <- c(
  "ID",
  "Item",
  "Original Component",
  "Key",
  "Primary Factor",
  "Primary Loading",
  "Secondary Factor",
  "Secondary Loading",
  "Weak Primary",
  "Cross-Loading"
)

knitr::kable(
  appendix_table,
  digits = 3
)
```

# Appendix B - Scoring Tables

Any raw score which results in a Z-score more extreme than 3 SD in either direction should be interpreted as standard score "<55" or ">145".

```{r}
#| echo: false
#| warning: false
#| message: false
#| results: asis

scoring_tables <- read_csv(
  "../output/barchard/barchard_scale_scoring_tables.csv",
  col_types = cols(
    scale = col_character(),
    raw_score = col_double(),
    standard_score = col_double(),
    percentile_rank = col_character(),
    ci_90 = col_character(),
    qualitative_descriptor = col_character()
  )
)

score_descriptives <- read_csv(
  "../output/barchard/barchard_scale_descriptives.csv",
  show_col_types = FALSE
)

score_scale_lookup <- unique(
  appendix_key[, c("scale_id", "scale")]
)

scoring_tables$scale <- score_scale_lookup$scale[
  match(scoring_tables$scale, score_scale_lookup$scale_id)
]

for (scale in score_scale_lookup$scale) {
  scale_id <- score_scale_lookup$scale_id[
    score_scale_lookup$scale == scale
  ]

  scale_table <- scoring_tables[
    scoring_tables$scale == scale,
    c(
      "raw_score",
      "standard_score",
      "percentile_rank",
      "ci_90",
      "qualitative_descriptor"
    )
  ]

  scale_desc <- score_descriptives[
    score_descriptives$scale == scale_id,
  ]

  colnames(scale_table) <- c(
    "Raw Score",
    "Standard Score",
    "Percentile Rank",
    "90% CI",
    "Qualitative Descriptor"
  )

  max_visible_raw <- max(scale_table$`Raw Score`)
  min_visible_raw <- min(scale_table$`Raw Score`)
  top_hidden_min <- max_visible_raw + 1
  top_hidden_max <- scale_desc$theoretical_max
  bottom_hidden_min <- scale_desc$theoretical_min
  bottom_hidden_max <- min_visible_raw - 1

  format_raw_range <- function(min_raw, max_raw) {
    if (min_raw > max_raw) {
      return(NA_character_)
    }

    if (min_raw == max_raw) {
      return(as.character(min_raw))
    }

    paste0(min_raw, "-", max_raw)
  }

  top_row <- if (!is.na(top_hidden_min) && top_hidden_min <= top_hidden_max) {
    data.frame(
      "Raw Score" = format_raw_range(top_hidden_min, top_hidden_max),
      "Standard Score" = ">145",
      "Percentile Rank" = ">99.9",
      "90% CI" = "-",
      "Qualitative Descriptor" = "Extremely High",
      check.names = FALSE
    )
  } else {
    NULL
  }

  bottom_row <- if (
    !is.na(bottom_hidden_max) &&
      bottom_hidden_min <= bottom_hidden_max
  ) {
    data.frame(
      "Raw Score" = format_raw_range(bottom_hidden_min, bottom_hidden_max),
      "Standard Score" = "<55",
      "Percentile Rank" = "<0.1",
      "90% CI" = "-",
      "Qualitative Descriptor" = "Extremely Low",
      check.names = FALSE
    )
  } else {
    NULL
  }

  scale_table[] <- lapply(scale_table, as.character)
  scale_table <- rbind(top_row, scale_table, bottom_row)

  cat(paste0("### ", scale, "\n"))
  print(knitr::kable(scale_table, align = c("r", "r", "r", "c", "l")))
  cat("\n\n")
}
```

# References
